---
title: "Introduction to Data Visualization with ggplot2"
author: "Andi"
Last updated: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo = TRUE, results='hide'}
# if we used both 'echo=TRUE' and 'results=hide' the pipe would not work properly
# if we used 'echo = FALSE' and 'results=hide' we would have only messages (i.e. attaching package) If we don't want them we set 'error = FALSE', 'warning = FALSE', and 'message = FALSE'.
library(dplyr)
library(ggplot2)
library(gridExtra) # display plots side by side
```

## Drawing your first plot

```{r}
# Load the ggplot2 package
library(ggplot2)

# Explore the mtcars data frame with str()
str(mtcars)

# Execute the following command
ggplot(mtcars, aes(cyl, mpg)) +
  geom_point()
```

## Data columns types affect plot types

```{r}
# Change the command below so that cyl is treated as factor
ggplot(mtcars, aes(factor(cyl), mpg)) +
  geom_point()
```

## The grammar of graphics

```{r}
# Edit to add a color aesthetic mapped to disp
ggplot(mtcars, aes(wt, mpg, color = disp)) +
  geom_point()

# Change the color aesthetic to a size aesthetic
ggplot(mtcars, aes(wt, mpg, size = disp)) +
  geom_point()
```

## ggplot2 layers

```{r}
load('diamonds.RData') # load R workspace

# Add geom_point() with +
ggplot(diamonds, aes(carat, price)) + geom_point()

# Add geom_smooth() with +
ggplot(diamonds, aes(carat, price)) +
  geom_point() + geom_smooth()
```

## Changing one geom or every geom

```{r}
# Map the color aesthetic to clarity
ggplot(diamonds, aes(carat, price, color = clarity)) +
  geom_point() +
  geom_smooth()

# Make the points 40% opaque
ggplot(diamonds, aes(carat, price, color = clarity)) +
  geom_point(alpha = 0.4) +
  geom_smooth()
```

## Saving plots as variables

```{r}
# Draw a ggplot
plt_price_vs_carat <- ggplot(diamonds, aes(carat, price)) + geom_point()
plt_price_vs_carat

# From previous step
plt_price_vs_carat <- ggplot(diamonds, aes(carat, price))

# Edit this to make points 20% opaque: plt_price_vs_carat_transparent
plt_price_vs_carat_transparent <- plt_price_vs_carat + geom_point(alpha = 0.2)

# See the plot
plt_price_vs_carat_transparent


# From previous step
plt_price_vs_carat <- ggplot(diamonds, aes(carat, price))

# Edit this to map color to clarity,
# Assign the updated plot to a new object
plt_price_vs_carat_by_clarity <- plt_price_vs_carat + geom_point(aes(color = clarity))

# See the plot
plt_price_vs_carat_by_clarity

# Which I believe gives the very same plot as the following line of code
tmp <- ggplot(diamonds, aes(carat, price, color = clarity)) + geom_point()

grid.arrange(plt_price_vs_carat_by_clarity, tmp, ncol = 2)

```

## All about aesthetics: color, shape and size

```{r}
mtcars$fcyl <- factor(mtcars$cyl) # cyl transformed into a factor

# Map x to mpg and y to fcyl
ggplot(mtcars, aes(mpg, fcyl)) +
  geom_point()

# Swap mpg and fcyl
ggplot(mtcars, aes(fcyl, mpg)) +
  geom_point()

# Map x to wt, y to mpg and color to fcyl
ggplot(mtcars, aes(wt, mpg, color = fcyl)) +
  geom_point()

ggplot(mtcars, aes(wt, mpg, color = fcyl)) +
  # Set the shape and size of the points
  geom_point(shape = 1, size = 4)
```

## All about aesthetics: color vs. fill

```{r}
# Map fcyl to fill
ggplot(mtcars, aes(wt, mpg, fill = fcyl)) +
  geom_point(shape = 1, size = 4)

ggplot(mtcars, aes(wt, mpg, fill = fcyl)) +
  # Change point shape; set alpha
  geom_point(shape = 21, size = 4, alpha = 0.6)

# Map color to fam
ggplot(mtcars, aes(wt, mpg, fill = fcyl, color = fam)) +
  geom_point(shape = 21, size = 4, alpha = 0.6)
```

Shapely coding! Notice that mapping a categorical variable onto fill doesn't change the colors, although a legend is generated! This is because the default shape for points only has a color attribute and not a fill attribute! Use fill when you have another shape (such as a bar), or when using a point that does have a fill and a color attribute, such as `shape = 21`, which is a circle with an outline. Any time you use a solid color, make sure to use alpha blending to account for over plotting.

